A Simplified Formulation for Molecular Dynamics with Rigid Bodies

### 2016-11-01 rigid_bodies.tex
Order inversion of symplectic integration and exact rotation

Eq.~\ref{eq:EDO_system} cannot be integrated analytically for a general potential $U(\vt r,\vt q)$, but closed-form solutions for Eqs.~\ref{eq:EDO_q} and \ref{eq:EDO_pi} are known in the case when $\vt \tau = \vt 0$. Torque-free rotations are useful in Molecular Dynamics as part of numerical integration methods based on splitting strategies. This is the topic addressed in Sec.~\ref{sec:symplectic}. As shown in Appendix~\ref{sec:derivation_euler}, Eq.~\ref{eq:EDO_pi} results in the well-known Euler equation $\mt I \dot{\vt \omega} + \vt \omega \times (\mt I \vt \omega) = \mt A \vt \tau$, which is totally decoupled from Eq.~\ref{eq:EDO_q} when $\vt \tau = \vt 0$. It is a known fact that both the rotational energy $K_r = \frac{1}{2}\tr{\vt \omega}{\mt I}{\vt \omega}$ and the space-fixed angular momentum vector $\vt L$ (whose squared norm is $L^2 = \tr{\vt \omega}{\mt I}^2{\vt \omega}$) are integrals of motion in a free rotation. This allows us to solve the Euler equation analytically for $\vt \omega$. In the case of a spherical top ($I_1 = I_2 = I_3$), the solution is simply $\vt \omega(t) = \vt \omega(0)$. Inspired by the derivations in Refs.~\onlinecite{Landau1976, Thorn2012, vanZon2007}, we present the general solution in a form that facilitates computer implementation and allows a unified treatment for both asymmetric ($I_1 > I_2 > I_3$) and symmetric tops (either $I_1 = I_2$ or $I_2 = I_3$) if infinities are properly dealt with (IEEE 754 standard). From an initial condition $\vt \omega(0) = \vt \omega_0 = \tr{[\omega_1^0 \; \omega_2^0 \; \omega_3^0]}$, the solution starts by computing $K_r = \frac{1}{2}\tr{\vt \omega_0}{\mt I}{\vt \omega}_0$ and $L = (\tr{\vt \omega_0}{\mt I}^2{\vt \omega_0})^{1/2}$, followed by evaluating a set of prefactors defined as
\begin{align*}
a_1 &= \text{sgn}(\omega_1^0)\sqrt{\frac{L^2 - 2 I_3 K_r}{I_1(I_1 - I_3)}}, \\
a_2 &= \sqrt{\min(\lambda_1,\lambda_3)}, \quad \text{and} \\
a_3 &= \text{sgn}(\omega_3^0)\sqrt{\frac{2 I_1 K_r - L^2}{I_1(I_1 - I_3)}}, \\
\end{align*}
where $\lambda_1 = \frac{L^2 - 2 I_3 K_r}{I_2(I_2 - I_3)}$ and $\lambda_3 = \frac{2 I_1 K_r - L^2}{I_2(I_1 - I_2)}$. Note that the value of $a_2$ depends on a comparison between $\lambda_1$ and $\lambda_3$. This also occurs for $\vt \omega$, whose functional form is
\begin{subequations}
\begin{align}
\vt \omega(t) &= \begin{smallarray}{c}
a_1\text{cn}(\omega_p t + \epsilon) \\
a_2\text{sn}(\omega_p t + \epsilon) \\
a_3\text{dn}(\omega_p t + \epsilon)
\end{smallarray} \quad \text{if} \quad \lambda_1 < \lambda_3 \quad \text{or} \\
\vt \omega(t) &= \begin{smallarray}{c}
a_1\text{dn}(\omega_p t + \epsilon) \\
a_2\text{sn}(\omega_p t + \epsilon) \\
a_3\text{cn}(\omega_p t + \epsilon)
\end{smallarray} \quad \text{if} \quad \lambda_1 > \lambda_3,
\end{align}
\end{subequations}
where $\text{sn}$, $\text{cn}$, and $\text{dn}$ are the Jacobi elliptic functions\cite{Byrd_1971} with parameter $m = k^2 = {a_2}/{\text{max}(\lambda_1,\lambda_3)}$ and the constants $\omega_p$ and $\epsilon$ are given by
\begin{equation*}
\omega_p = -\frac{(I_1 - I_3)a_1 a_3}{I_2 a_2} \quad \text{and} \quad
\epsilon = \text{sn}^{-1}\left(\frac{\omega_2^0}{a_2}\right).
\end{equation*}

@@

>>>>

@@

Eq.~\ref{eq:EDO_system} cannot be integrated analytically for a general potential $U(\vt r,\vt q)$, but closed-form solutions for Eqs.~\ref{eq:EDO_q} and \ref{eq:EDO_pi} are known in the case when $\vt \tau = \vt 0$. Torque-free rotations are useful in Molecular Dynamics as part of numerical integration methods based on splitting strategies. This is the topic addressed in Sec.~\ref{sec:symplectic}. As shown in Appendix~\ref{sec:derivation_euler}, Eq.~\ref{eq:EDO_pi} results in the well-known Euler equation $\mt I \dot{\vt \omega} + \vt \omega \times (\mt I \vt \omega) = \mt A \vt \tau$, which is totally decoupled from Eq.~\ref{eq:EDO_q} when $\vt \tau = \vt 0$. It is a known fact that both the rotational energy $K_r = \frac{1}{2}\tr{\vt \omega}{\mt I}{\vt \omega}$ and the space-fixed angular momentum vector $\vt L$ (whose squared norm is $L^2 = \tr{\vt \omega}{\mt I}^2{\vt \omega}$) are integrals of motion in a free rotation. This allows us to solve the Euler equation analytically for $\vt \omega$. In the case of a spherical top ($I_1 = I_2 = I_3$), the solution is simply $\vt \omega(t) = \vt \omega(0)$. Inspired by the derivations in Refs.~\onlinecite{Landau1976, Thorn2012, vanZon2007}, we present the general solution in a form that facilitates computer implementation and allows a unified treatment for both asymmetric ($I_1 > I_2 > I_3$) and symmetric tops (either $I_1 = I_2$ or $I_2 = I_3$) if infinities are properly dealt with (IEEE 754 standard). From an initial condition $\vt \omega(0) = \vt \omega_0 = \tr{[\omega_1^0 \; \omega_2^0 \; \omega_3^0]}$, the solution starts by computing $K_r = \frac{1}{2}\tr{\vt \omega_0}{\mt I}{\vt \omega}_0$ and $L = (\tr{\vt \omega_0}{\mt I}^2{\vt \omega_0})^{1/2}$, followed by evaluating a set of prefactors defined as
\begin{align*}
a_1 &= \text{sgn}(\omega_1^0)\sqrt{\frac{L^2 - 2 I_3 K_r}{I_1(I_1 - I_3)}}, \\
a_2 &= \sqrt{\min(\lambda_1,\lambda_3)}, \quad \text{and} \\
a_3 &= \text{sgn}(\omega_3^0)\sqrt{\frac{2 I_1 K_r - L^2}{I_1(I_1 - I_3)}}, \\
\end{align*}
where $\lambda_1 = \frac{L^2 - 2 I_3 K_r}{I_2(I_2 - I_3)}$ and $\lambda_3 = \frac{2 I_1 K_r - L^2}{I_2(I_1 - I_2)}$. Note that the value of $a_2$ depends on a comparison between $\lambda_1$ and $\lambda_3$. This also occurs for $\vt \omega$, whose functional form is
\begin{subequations}
\begin{align}
\vt \omega(t) &= \left[\begin{array}{c}
a_1\text{cn}(\omega_p t + \epsilon) \\
a_2\text{sn}(\omega_p t + \epsilon) \\
a_3\text{dn}(\omega_p t + \epsilon)
\end{array}\right] \quad \text{if} \quad \lambda_1 \le \lambda_3 \quad \text{or} \\
\vt \omega(t) &= \left[\begin{array}{c}
a_1\text{dn}(\omega_p t + \epsilon) \\
a_2\text{sn}(\omega_p t + \epsilon) \\
a_3\text{cn}(\omega_p t + \epsilon)
\end{array}\right] \quad \text{if} \quad \lambda_1 > \lambda_3,
\end{align}
\end{subequations}
where $\text{sn}$, $\text{cn}$, and $\text{dn}$ are the Jacobi elliptic functions\cite{Byrd_1971} with parameter $m = k^2 = {a_2}/{\text{max}(\lambda_1,\lambda_3)}$ and the constants $\omega_p$ and $\epsilon$ are given by
\begin{equation*}
\omega_p = \frac{(I_3 - I_1)a_1 a_3}{I_2 a_2} \quad \text{and} \quad
\epsilon = \text{sn}^{-1}\left(\frac{\omega_2^0}{a_2}\right).
\end{equation*}

In the equations above, $\eta = \frac{\alpha^2}{1-\alpha^2}$, the function $\text{sgn}(\cdot)$ returns $-1$ if its argument is negative or $+1$ otherwise, the operator $\lfloor \cdot \rceil$ rounds a real value to its nearest integer (with the unusual feature that halfway values are rounded towards zero), and the function $\Theta$ is defined in terms of the Carlson elliptic integral $R_J$ as
\begin{equation*}
\Theta(x,n) = -\frac{1}{3}nx^3 R_J(1-x^2,1-mx^2,1,1+nx^2).
\end{equation*}
